\LaTeX\ Author Guidelines for ICCV Proceedings

### 2021-03-12 chapters/3_algorithm.tex
Comments from Fred; move prop to appendix

\textbf{Multicut objective} -- We call the set $\Pi=\{S_1,\ldots,S_K\}$ a \emph{clustering} or \emph{partitioning} if $V = \cup_{S\in\Pi} S $, $\,S \cap S' = \emptyset$ for different clusters $S, S'\in \Pi$ and every cluster $S \in \Pi$ induces a connected subgraph of $\mathcal{G}$. 
For any clustering $\Pi$ of $\mathcal{G}$, we denote as $E^0_\Pi= \{ e_{uv} \in E \,|\, \exists S \in \Pi : u,v \in S \}$ the set of edges linking nodes in the same cluster. Its complementary set $E_\Pi^1= E \setminus E^0_\Pi$ of edges linking nodes belonging to distinct clusters, is known as the \emph{multicut} of $\mathcal{G}$ associated to clustering $\Pi$. The instance of the NP-hard \emph{minimum cost multicut problem} w.r.t. $\mathcal{G}(V,E,w_e)$ is the task of finding a clustering that optimally balance the attraction and repulsion in the graph and is given by the following binary integer program:
% \begin{equation}
% E_\Pi^0 \equiv \{ e_{uv} \in E \,|\, \exists S \in \Pi : u \in S \, \text{and} \, v \in S \}, \qquad E^1_\Pi \equiv E \setminus E^0_\Pi.
% \end{equation}
% \begin{align}
% E_\Pi^0 &= \{ e_{uv} \in E \,|\, \exists S \in \Pi : u \in S \, \text{and} \, v \in S \}, \\
% E^1_\Pi &= E \setminus E^0_\Pi.
% \end{align}
\begin{equation}\label{eq:MC_objective}
% \min_\Pi \texttt{MC}(\Pi) \equiv
 \min_\Pi \sum_{e\in E} \cost_e x_e^\Pi,  \qquad \text{where} \quad x^\Pi_e = 
 \begin{cases} 
 1 & \text{if } e\in E^1_\Pi \\
 0 & \text{otherwise}.
 \end{cases}
\end{equation}
% In the next chapters we will use this objective as a way to measure how balanced is a clustering found by the tested agglomerative clustering algorithms.

\textbf{Linkage criteria and hierarchical trees} -- 
% We also denote as $S_u$ the cluster associated with node $u$ \TODO{Needed?}.
% We call two clusters $S,S'$ \emph{adjacent} if there exists at least one edge ${e_{ts}\in E}$ connecting a node $t\in S_u$ to a node $s\in S_v$. 
Let the interaction $\interact(S,S')$ between two clusters $S,S'$ be defined as a function $\interact{}:2^{V} \times 2^{V} \rightarrow \mathbb{R}$, named \emph{linkage criterion}, depending on the weights of \emph{all} edges connecting clusters $S$ and $S'$, i.e. ${(S \times S') \cap E}$. 
The linkage criteria tested in this article are listed and defined in Table \ref{tab:linkage-criteria}.
A \emph{dendrogram} $T$ is a rooted binary tree\footnote{In general, one could look at trees that are not binary. However, the algorithms discussed in this paper always generate binary hierarchical trees, so nothing would be gained by this generalization.} representing the merging order of an agglomerative algorithm, such that the leaves of the tree are in one-to-one correspondence with $V$. 
Let $T_{\mathrm{R}},T_{\mathrm{L}}\subset T$ denote the subtrees rooted at the two children of the root node in $T$.
For any two leaves $u,v \in V$, let $T[u \vee v]$ be the subtree rooted at the least common ancestor $(u \vee v)\in T$ of nodes $u$ and $v$ (furthest from the root), and let \texttt{leaves}$(T[u \vee v])\subseteq V$ be the set of leaves of this subtree. 
% \TODO{Here we talk about the algorithm already, so we could move it to next section and the flow could be better. But we need a tree in the alg. So we better do this later}
% Algorithm \ref{main_alg} returns a binary tree denoted by $T^*$.
Given an agglomerative algorithm with merging tree $T$, let $h_T:T \rightarrow \mathbb{N}$ denote the \emph{dendrogram-height} of each $(u\vee v)\in T$, which is defined as the iteration number at which nodes $u,v\in V$ were merged by the algorithm (see example in Fig.~\ref{fig:intro_figure}b). We also define $\treeHeight(u,v)$ as the signed interaction $\interact{}(S,S')$ between the two clusters $S,S'$ that were merged at iteration $h_T(u\vee v)$: 
% Then, given a binary tree $T$ and a linkage criterion $\interact$, we can associate a \emph{signed similarity} function $\treeHeight$ to each pair of distinct original nodes $u\neq v$, with $u, v \in V$:
\begin{equation}\label{eq:def_dendr_interact}
\treeHeight(u,v) \equiv \interact{} \big( \text{\texttt{leaves}}(T_{\mathrm{R}}[u \vee v]), \text{\texttt{leaves}}(T_{\mathrm{L}}[u \vee v]) \big)
\end{equation}
% \OPTIONAL{In other words, $h_T(u,v)$ and $\treeHeight(u,v)$ represent the \emph{dendrogram-height} and \emph{signed-interaction} associated to each $u \vee v$ node in the tree $T$ (see toy example in \TODO{Fig.~1}).}

@@

In \emph{Phase~One}, \algname{} selects the pair of clusters with the highest absolute interaction $|\interact(S, S')|$, so that the most attractive and the most repulsive pairs are analyzed first. If the interaction is repulsive and the algorithm option \emph{addCannotLinkConstraints} is \texttt{True}, then the two clusters are constrained so that their members can never merge in subsequent steps of \emph{Phase~One}. If the interaction is attractive, then the clusters are merged, provided that they were not previously constrained. 
After each merge, the interaction between the merged cluster and its neighbors is updated according to one of the linkage criteria $\interact(S, S')$ listed in Table \ref{tab:linkage-criteria}. \emph{Phase~One} terminates when all the remaining clusters are either constrained or share repulsive interactions.

\emph{Phase Two:} Now that the clusters have grown in size, the algorithm removes the constraints previously introduced in \emph{Phase~One} and merges all the clusters that still share an attractive interaction, merging the most attractive one first\footnote{Note that in the version of \algname{} with \emph{addCannotLinkConstraints} set to \texttt{False}, nothing happens in \emph{Phase~Two}.}. The final clustering $\Pi^*$ returned by \algname{} is found at the end of \emph{Phase~Two} and it is then composed of clusters sharing only mutual repulsive interactions.

Finally, in \emph{Phase~3}, the algorithm keeps merging all clusters until only a single one is left and then returns the hierarchical tree $T^*$ representing the full sequence of merging steps.
The algorithm was implemented using a standard HC implementation with computational complexity $\mathcal{O}(N^2 \log N)$ (see Appendix \ref{sec:detailed_impl} for more details).

@@

\begin{algorithm}[t]
\footnotesize
  \begin{flushleft}
  \footnotesize
  \caption{\algname{}}
  % \caption{\algname{}: generalized algorithm for signed graph partitioning}
   \hspace*{\algorithmicindent} \textbf{Input:} Graph $\mathcal{G}(V,E,w^+,w^-)$; linkage criterion $\interact{}$; \\ 
   \hspace*{4.3em}boolean {\color{blue}addCannotLinkConstraints}  \\
  \hspace*{\algorithmicindent} \textbf{Output:} Final clustering $\Pi^*$, rooted binary hierarchical tree $T^*$\\
  \hspace*{\algorithmicindent} 
  \begin{algorithmic}[1]
  \footnotesize
  % \small
      \State Initial clustering: $\Pi=\{\{v_1\}, \ldots, \{v_{|V|}\}\}$
      \State Initialize hierarchical tree $T$ with leaves nodes $V=\{v_1,\ldots,v_{|V|}\}$
      \State Initialize interactions between clusters with $\cost_e = w^+_e - w^-_e$
      \State \emph{// Phase 1: Merge positive interactions (possibly using constraints)}
      \State Push absolute interactions $|w_e|$ to priority queue $PQ$
      \Repeat 
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $|\interact{}(S,S')|$
        \If{\big[{\color{green}\textbf{$\interact{}(S,S') > 0$}}\big] \textbf{and} \big[$S,S'$ \textbf{not} constrained\big]}
          \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
          \State Update interactions \& constraints with neighboring clusters
        \ElsIf{{\color{blue}addCannotLinkConstr} \textbf{and}  \big[{\color{red}\textbf{$\interact{}(S,S') \leq 0$}}\big]}
          \State Add CannotLink Constraint between $S$ and $S'$
        \EndIf
      \Until{\big[$PQ$ is empty\big]}
      % \State
      % \If{{\color{blue}addCannotLinkConstr}} % \Comment{Step 2: release constraints}
      \State \emph{// Phase 2: Remove constraints \& merge all positive interactions}
      \State Push signed interactions $\interact{}(S,S')$ to PQ, $\forall S, S' \in \Pi$
      \Repeat 
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $\interact{}(S,S')$
        \If{\big[{\color{green}\textbf{$\interact{}(S,S') > 0$}}\big]}
          \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
          \State Update interactions with neighboring clusters
        \EndIf
      \Until{\big[{\color{red}\textbf{$\interact{}(S,S') \leq 0$}}\big]} 
      \State Save the final clustering $\Pi^* \gets \Pi$ 
      \State \emph{// Phase 3: Merge negative interactions until one single cluster is left}
      \Repeat
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $\interact{}(S,S')$
        \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
        \State Update interactions with neighboring clusters
      \Until{\big[Only one cluster is left in $\Pi$\big]} 
      % \EndIf
      \State
      \Return $\Pi^*$, $T^*\gets T$
  \end{algorithmic}
    \label{main_alg}
  \end{flushleft}

@@

\paragraph{Sum Linkage} -- 
On signed graphs, the sum of two attractive (or repulsive) interactions is still attractive (repulsive). On the other hand, on positive weighted graphs, a strong attractive interaction could be obtained by summing many weak interactions, which is often not desirable. This explains why, to our knowledge, an agglomerative algorithm with sum linkage has never been used on positive weighted graphs. On signed graphs, such an algorithm was pioneered in \cite{levinkov2017comparative,keuper2015efficient} and was named Greedy Agglomerative Edge Contraction (GAEC)\footnote{An algorithm equivalent to GAEC was recently independently re-proposed in \cite{chehreghani2020hierarchical}.}. All variations of \algname{} decrease the multicut objective defined in Eq.~\ref{eq:MC_objective} each time two clusters with positive interaction are merged. But only GAEC always makes the \emph{greedy-choice} that will mostly decrease the multicut objective at every step. The authors of \cite{levinkov2017comparative} also proposed a version of GAEC with cannot-link constraints named \emph{GreedyFixation}, but the constrained sum-link version of \algname{} proposed here (named \emph{HCC-Sum} in Table~\ref{tab:linkage-criteria}) did perform better in our experiments.

\paragraph{AbsMax Linkage} -- This linkage method is also specific to signed graphs, since on positive-weighted graphs it would be equivalent to single linkage. Here, we prove that the Mutex Watershed Algorithm \cite{wolf2018mutex} can be seen as an agglomerative algorithm with AbsMax linkage (proofs of the following propositions are given in Appendix \ref{sec:proposition_proofs}):
\begin{restatable}{prop}{absmaxmutex}
\label{prop:absmax_mutex}
The \algname{} Algorithm \ref{main_alg} with AbsMax linkage (see def. in Table \ref{tab:linkage-criteria}) returns the same final clustering $\Pi^*_{\mathrm{AbsMax}}$ whether or not cannot-link constraints are introduced. In particular, $\Pi^*_{\mathrm{AbsMax}}$ is equal to the clustering returned by the Mutex Watershed Algorithm (MWS) \cite{wolf2018mutex}, which has empirical complexity $\mathcal{O}(N \log N)$. 
\end{restatable}
% \begin{prop} 
% \end{prop}

\paragraph{Average, Single, and Complete Linkage} -- These three linkage criteria have been throughly studied on positive-weighted graphs, but never on signed graphs. In the recent related work \cite{chehreghani2020hierarchical}, the authors prove that applying these three linkage methods to a signed graph is equivalent to applying them to the positive-weighted graph obtained by shifting all edge weights by a constant.  Here, we prove which of the algorithms included in our generalized framework have this invariance-property:
\begin{restatable}{prop}{invariantAlgs}
\label{prop:weight_shift_invariant}
We call an agglomerative algorithm ``weight-shift invariant'' if the structure of the merging tree $T$ returned by such algorithm is invariant w.r.t. a shift of all edge weights $w_e$  by a constant $\alpha\in \mathbb{R}$. Among the versions of \algname{} studied in this paper, only hierarchical clustering using Average (HC-Avg), Single (HC-Single), and Complete linkage (HC-Complete) are weight-shift-invariant (see algorithms included in the green box in Table~\ref{tab:linkage-criteria}).
\end{restatable}
\noindent To our knowledge, Average and Single linkage methods have never been combined with cannot-link constraints on signed graphs, so the constrained hierarchical clustering algorithms that in Table~\ref{tab:linkage-criteria} are named \emph{HCC-Avg} and \emph{HCC-Single} are new. In Lemma \ref{lemma:absMax_and_complete_property} (Appendix \ref{sec:proposition_proofs}), we prove that with a Complete linkage the agglomeration order does not change whether constraints are introduced or not.

@@

\paragraph{Algorithms defining an ultrametric} -- The connection between agglomerative algorithms and ultrametrics\footnote{A metric space $(X,d)$ is an \emph{ultrametric} if for every $x,y,z \in X$, $d(x,y)\leq \max \{d(x,z), d(y,z)\}$.} is well known \cite{johnson1967hierarchical,milligan1979ultrametric}. Usually, ultrametrics are associated to strictly positive \emph{similarity} or \emph{dissimilarity} measures on a graph. Here, we generalize this concept to signed graphs. To do so, we first map the signed interaction $\treeHeight{}$ defined in Eq.~\ref{eq:def_dendr_interact} to positive ``pseudo-distances'' $d_{T}:V \times V \rightarrow \mathbb{R}^{+}$:
\begin{align}\label{eq:UM_def}
d_{T}(u,v)\equiv &\begin{cases}
0 & \text{if}\,\, u=v\\
M-\treeHeight(u,v) & \text{if}\,\, u\neq v
\end{cases}\quad \forall u,v\in V\\
\mathrm{where}& \quad  M \equiv  \epsilon + \max_{u',v'\in V,\,u'\neq v'}\treeHeight(u',v')
\end{align}
and where $\epsilon > 0$. By definition, we have that $d_{T}(u,v)=d_{T}(v,u)$ and $d_{T}(u,u)= 0$. We then prove the following two propositions:
\begin{restatable}{prop}{firstUltraMetricProperty}
\label{prop:ultraMetric1}
Consider a graph $\mathcal{G}(V,E,w_e)$, a linkage criterion $\interact{}$, and an agglomerative algorithm returning a binary rooted tree $T$ with height $h_T$. Then, $(V, d_{T})$ defined in Eq.~\ref{eq:UM_def} is an ultrametric if and only if the following is true
 % iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than the interaction of any other $(u\vee v)$-descendant node and higher than the interaction of any other $(u\vee v)$-ancestor node in the tree $T$. In formulae, this is equivalent to saying that, $\forall u,v,t \in V$:
\begin{equation}\label{eq:UM_assumption}
 h_T(u \vee v) < h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) \geq \interact{}_{T}(u,t),
 % h_T(u \vee v) < h_T(u \vee t) \Leftrightarrow \interact{}_{T}(u,v) > \interact{}_{T}(u,t). 
\end{equation}
which in words means: if the algorithm merges nodes $u,v$ before to merge nodes $u,t$, then the signed interaction $\interact{}_{T}(u,v)$ between $u$ and $v$ has to be higher or equal than $\interact{}_{T}(u,t)$.
% In words, $(V, d_{T})$ is an ultrametric iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than any other $(u\vee v)$-descendant node and higher than any other $(u\vee v)$-ancestor node in the tree $T$.
\end{restatable}
\begin{restatable}{prop}{secondUltraMetricProperty}
\label{prop:ultraMetric2}
% Consider the binary rooted tree $T^*$ returned by the \TODO{all variants in Table 1} \algname{} Algorithm~\ref{main_alg}, given a choice of linkage criteria $\interact{}$ and the use (or not) of CannotLinkConstraints. 
Among the algorithms included in the \algname{} framework (see Table \ref{tab:linkage-criteria}), only Mutex Watershed and hierarchical clustering with Average (HC-Avg), Single (HC-Single) and Complete linkage (HC-Complete) define an ultrametric $(V, d_{T^*})$, where $d_{T^*}$ is defined in Eq.~\ref{eq:UM_def} and $T^*$ is the tree returned by the \algname{} Algorithm~\ref{main_alg}.
\end{restatable}

>>>>

\textbf{Multicut objective} -- We call the set $\Pi=\{S_1,\ldots,S_K\}$ a \emph{clustering} or \emph{partitioning} if $V = \cup_{S\in\Pi} S $, $\,S \cap S' = \emptyset$ for different clusters $S, S'$ and every cluster $S \in \Pi$ induces a connected subgraph of $\mathcal{G}$. 
For any clustering $\Pi$ of $\mathcal{G}$, we denote as $E^0_\Pi= \{ e_{uv} \in E \,|\, \exists S \in \Pi : u,v \in S \}$ the set of edges linking nodes in the same cluster. Its complementary set $E_\Pi^1= E \setminus E^0_\Pi$ of edges linking nodes belonging to distinct clusters, is known as the \emph{multicut} of $\mathcal{G}$ associated to clustering $\Pi$. The instance of the NP-hard \emph{minimum cost multicut problem} w.r.t. $\mathcal{G}(V,E,w_e)$ is the task of finding a clustering that optimally balance the attraction and repulsion in the graph and is given by the following binary integer program:
% \begin{equation}
% E_\Pi^0 \equiv \{ e_{uv} \in E \,|\, \exists S \in \Pi : u \in S \, \text{and} \, v \in S \}, \qquad E^1_\Pi \equiv E \setminus E^0_\Pi.
% \end{equation}
% \begin{align}
% E_\Pi^0 &= \{ e_{uv} \in E \,|\, \exists S \in \Pi : u \in S \, \text{and} \, v \in S \}, \\
% E^1_\Pi &= E \setminus E^0_\Pi.
% \end{align}
\begin{equation}\label{eq:MC_objective}
% \min_\Pi \texttt{MC}(\Pi) \equiv
 \min_\Pi \sum_{e\in E} \cost_e x_e^\Pi,  \qquad \text{where} \quad x^\Pi_e = 
 \begin{cases} 
 1 & \text{if } e\in E^1_\Pi \\
 0 & \text{otherwise}.
 \end{cases}
\end{equation}
% In the next chapters we will use this objective as a way to measure how balanced is a clustering found by the tested agglomerative clustering algorithms.

\textbf{Linkage criteria and hierarchical trees} -- 
% We also denote as $S_u$ the cluster associated with node $u$ \TODO{Needed?}.
% We call two clusters $S,S'$ \emph{adjacent} if there exists at least one edge ${e_{ts}\in E}$ connecting a node $t\in S_u$ to a node $s\in S_v$. 
Let the interaction $\interact(S,S')$ between two clusters $S,S'$ be defined as a function $\interact{}:2^{V} \times 2^{V} \rightarrow \mathbb{R}$, named \emph{linkage criterion}, depending on the weights of \emph{all} edges connecting clusters $S$ and $S'$, i.e. ${(S \times S') \cap E}$. 
The linkage criteria tested in this article are listed and defined in Table \ref{tab:linkage-criteria}.
A \emph{dendrogram} $T$ is a rooted binary tree\footnote{In general, one could look at trees that are not binary. However, the algorithms discussed in this paper always generate binary hierarchical trees, so nothing would be gained by this generalization.} representing the merging order of an agglomerative algorithm, such that the leaves of the tree are in one-to-one correspondence with $V$. 
Let $T_{\mathrm{R}},T_{\mathrm{L}}\subset T$ denote the subtrees rooted at the two children of the root node in $T$.
For any two leaves $u,v \in V$, let $T[u \vee v]$ be the subtree rooted at the least common ancestor $(u \vee v)\in T$ of nodes $u$ and $v$ (furthest from the root), and let \texttt{leaves}$(T[u \vee v])\subseteq V$ be the set of leaves of this subtree. 
% \TODO{Here we talk about the algorithm already, so we could move it to next section and the flow could be better. But we need a tree in the alg. So we better do this later}
% Algorithm \ref{main_alg} returns a binary tree denoted by $T^*$.
Given an agglomerative algorithm with merging tree $T$, let $h_T:T \rightarrow \mathbb{N}$ denote the \emph{dendrogram-height} of each $(u\vee v)\in T$, which is defined as the iteration number at which nodes $u,v\in V$ were merged by the algorithm (see example in Fig.~\ref{fig:intro_figure}a). We also define $\treeHeight(u,v)$ as the signed interaction $\interact{}(S,S')$ between the two clusters $S,S'$ that were merged at iteration $h_T(u\vee v)$: 
% Then, given a binary tree $T$ and a linkage criterion $\interact$, we can associate a \emph{signed similarity} function $\treeHeight$ to each pair of distinct original nodes $u\neq v$, with $u, v \in V$:
\begin{equation}\label{eq:def_dendr_interact}
\treeHeight(u,v) \equiv \interact{} \big( \text{\texttt{leaves}}(T_{\mathrm{R}}[u \vee v]), \text{\texttt{leaves}}(T_{\mathrm{L}}[u \vee v]) \big)
\end{equation}
% \OPTIONAL{In other words, $h_T(u,v)$ and $\treeHeight(u,v)$ represent the \emph{dendrogram-height} and \emph{signed-interaction} associated to each $u \vee v$ node in the tree $T$ (see toy example in \TODO{Fig.~1}).}

@@

In \emph{Phase~One}, \algname{} selects the pair of clusters with the highest absolute interaction $|\interact(S, S')|$, so that the most attractive and the most repulsive pairs are analyzed first. If the interaction is repulsive and the algorithm option \emph{addCannotLinkConstraints} is \texttt{True}, then the two clusters are constrained so that their members can never merge in subsequent steps of \emph{Phase~One}. If the interaction is attractive, then the clusters are merged, provided that they were not previously constrained. 
After each merge, the interaction between the merged cluster and its neighbors is updated according to one of the linkage criteria $\interact(S, S')$ listed in Table \ref{tab:linkage-criteria}. \emph{Phase~One} terminates when all the remaining clusters are either constrained or share repulsive interactions. Note that, on unsigned graphs, in \emph{Phase 1} \algname{} merges all nodes in a single cluster and is then equivalent to a standard hierarchical clustering algorithm.

\emph{Phase Two:} Now that the clusters have grown in size, the algorithm removes the constraints previously introduced in \emph{Phase~One} and merges all the clusters that still share an attractive interaction, merging the most attractive one first\footnote{Note that in the version of \algname{} without \emph{cannotLinkConstraints}, nothing happens in \emph{Phase~Two} because all remaining interactions are repulsive.}. The final clustering $\Pi^*$ returned by \algname{} is found at the end of \emph{Phase~Two} and it is then composed of clusters sharing only mutual repulsive interactions.

Finally, in \emph{Phase~3}, the algorithm keeps merging all clusters until only a single one is left and then returns the hierarchical tree $T^*$ representing the full sequence of merging steps.
The algorithm was implemented using a standard HC implementation with computational complexity $\mathcal{O}(N^2 \log N)$ (details left in Appendix \ref{sec:detailed_impl}).

@@

\begin{algorithm}[t]
\footnotesize
  \begin{flushleft}
  \footnotesize
  \caption{\algname{}}
  % \caption{\algname{}: generalized algorithm for signed graph partitioning}
   \hspace*{\algorithmicindent} \textbf{Input:} Graph $\mathcal{G}(V,E,w^+,w^-)$; linkage criterion $\interact{}$; \\ 
   \hspace*{4.3em}boolean {\color{blue}addCannotLinkConstraints}  \\
  \hspace*{\algorithmicindent} \textbf{Output:} Final clustering $\Pi^*$, rooted binary hierarchical tree $T^*$\\
  \hspace*{\algorithmicindent} 
  \begin{algorithmic}[1]
  \footnotesize
  % \small
      \State Initial clustering: $\Pi=\{\{v_1\}, \ldots, \{v_{|V|}\}\}$
      \State Initialize hierarchical tree $T$ with leaf nodes  $V=\{v_1,\ldots,v_{|V|}\}$
      \State Initialize interactions between clusters with $\cost_e = w^+_e - w^-_e$
      \State \emph{// Phase 1: Merge positive interactions (possibly using constraints)}
      \State Push absolute interactions $|w_e|$ to priority queue $PQ$
      \Repeat 
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $|\interact{}(S,S')|$
        \If{\big[{\color{green}\textbf{$\interact{}(S,S') > 0$}}\big] \textbf{and} \big[$S,S'$ \textbf{not} constrained\big]}
          \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
          \State Update interactions \& constraints with neighboring clusters
        \ElsIf{{\color{blue}addCannotLinkConstr} \textbf{and}  \big[{\color{red}\textbf{$\interact{}(S,S') \leq 0$}}\big]}
          \State Add CannotLink Constraint between $S$ and $S'$
        \EndIf
      \Until{\big[$PQ$ is empty\big]}
      % \State
      % \If{{\color{blue}addCannotLinkConstr}} % \Comment{Step 2: release constraints}
      \State \emph{// Phase 2: Remove constraints \& merge all positive interactions}
      \State Push signed interactions $\interact{}(S,S')$ to PQ, $\forall S, S' \in \Pi$
      \Repeat 
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $\interact{}(S,S')$
        \If{\big[{\color{green}\textbf{$\interact{}(S,S') > 0$}}\big]}
          \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
          \State Update interactions with neighboring clusters
        \EndIf
      \Until{\big[{\color{red}\textbf{$\interact{}(S,S') \leq 0$}}\big]} 
      \State Save the final clustering $\Pi^* \gets \Pi$ 
      \State \emph{// Phase 3: Merge negative interactions until one single cluster is left}
      \Repeat
        \State Pop $S,S'\in\Pi$ from $PQ$ with highest interaction $\interact{}(S,S')$
        \State Merge clusters $S$, $S'$ and update hierarchical tree $T$
        \State Update interactions with neighboring clusters
      \Until{\big[Only one cluster is left in $\Pi$\big]} 
      % \EndIf
      \State
      \Return $\Pi^*$, $T^*\gets T$
  \end{algorithmic}
    \label{main_alg}
  \end{flushleft}

@@

\paragraph{Sum Linkage} -- 
On signed graphs, the sum of two attractive (or repulsive) interactions is still attractive (repulsive). On the other hand, on unsigned graphs, a strong attractive interaction could be obtained by summing many weak interactions, which depending on the application could be undesirable. This explains why, to our knowledge, an agglomerative algorithm with sum linkage has never been used on unsigned graphs. On signed graphs, such an algorithm was pioneered in \cite{levinkov2017comparative,keuper2015efficient} and was named Greedy Agglomerative Edge Contraction (GAEC)\footnote{An algorithm equivalent to GAEC was recently independently re-proposed in \cite{chehreghani2020hierarchical}.}. All variations of \algname{} decrease the multicut objective defined in Eq.~\ref{eq:MC_objective} each time two clusters with positive interaction are merged. But only GAEC always makes the \emph{greedy-choice} that mostly decreases the multicut objective at each iteration. The authors of \cite{levinkov2017comparative} also proposed a version of GAEC with cannot-link constraints named \emph{GreedyFixation}, but the constrained sum-link version of \algname{} proposed here (named \emph{HCC-Sum} in Table~\ref{tab:linkage-criteria}) did perform better in our experiments.

\paragraph{AbsMax Linkage} -- This linkage method is also specific to signed graphs, since on unsigned graphs it would be equivalent to single linkage. Here, we prove that the Mutex Watershed Algorithm \cite{wolf2018mutex} can be seen as an agglomerative algorithm with AbsMax linkage (proofs of the following three propositions are given in Appendix \ref{sec:proposition_proofs}):
\begin{restatable}{prop}{absmaxmutex}
\label{prop:absmax_mutex}
The \algname{} Algorithm \ref{main_alg} with AbsMax linkage (see def. in Table \ref{tab:linkage-criteria}) returns the same final clustering $\Pi^*_{\mathrm{AbsMax}}$ whether or not cannot-link constraints are introduced. In particular, $\Pi^*_{\mathrm{AbsMax}}$ is equal to the clustering returned by the Mutex Watershed Algorithm (MWS) \cite{wolf2018mutex}, which has empirical complexity $\mathcal{O}(N \log N)$. 
\end{restatable}
% \begin{prop} 
% \end{prop}

\paragraph{Average, Single, and Complete Linkage} -- These three linkage criteria have been throughly studied on unsigned graphs, but never on signed graphs. In the recent related work \cite{chehreghani2020hierarchical}, the authors prove that applying these three linkage methods to a signed graph is equivalent to applying them to the unsigned graph obtained by shifting all edge weights by a constant.  Here, we prove which of the algorithms included in our generalized framework have this invariance-property:
\begin{restatable}{prop}{invariantAlgs}
\label{prop:weight_shift_invariant}
We call an agglomerative algorithm ``weight-shift invariant'' if the structure of the merging tree $T$ returned by such algorithm is invariant w.r.t. a shift of all edge weights $w_e$  by a constant $\alpha\in \mathbb{R}$. Among the versions of \algname{} studied in this paper, only hierarchical clustering using Average (HC-Avg), Single (HC-Single), and Complete linkage (HC-Complete) are weight-shift-invariant (see algorithms included in the green box in Table~\ref{tab:linkage-criteria}).
\end{restatable}
\noindent To our knowledge, Average and Single linkage methods have never been combined with cannot-link constraints on signed graphs, so the constrained hierarchical clustering algorithms \emph{HCC-Avg} and \emph{HCC-Single} are new (see Table~\ref{tab:linkage-criteria}). In Lemma \ref{lemma:absMax_and_complete_property} (Appendix \ref{sec:proposition_proofs}), we prove that with a Complete linkage the agglomeration order does not change whether constraints are introduced or not.

@@

\paragraph{Algorithms defining an ultrametric} -- The connection between agglomerative algorithms and ultrametrics\footnote{A metric space $(X,d)$ is an \emph{ultrametric} if for every $x,y,z \in X$, $d(x,y)\leq \max \{d(x,z), d(y,z)\}$.} is well known \cite{johnson1967hierarchical,milligan1979ultrametric}. Usually, ultrametrics are associated to strictly positive \emph{similarity} or \emph{dissimilarity} measures on a graph. Here, we generalize this concept to signed graphs. To do so, we first map the signed interaction $\treeHeight{}$ defined in Eq.~\ref{eq:def_dendr_interact} to positive ``pseudo-distances'' $d_{T}:V \times V \rightarrow \mathbb{R}^{+}$:
\begin{align}\label{eq:UM_def}
d_{T}(u,v)\equiv &\begin{cases}
0 & \text{if}\,\, u=v\\
M-\treeHeight(u,v) & \text{if}\,\, u\neq v
\end{cases}\quad \forall u,v\in V\\
\mathrm{where}& \quad  M \equiv  \epsilon + \max_{u',v'\in V,\,u'\neq v'}\treeHeight(u',v')
\end{align}
and where $\epsilon > 0$. 
% By definition, we have that $d_{T}(u,v)=d_{T}(v,u)$ and $d_{T}(u,u)= 0$. 
We then prove the following proposition:
\begin{restatable}{prop}{secondUltraMetricProperty}
\label{prop:ultraMetric2}
% Consider the binary rooted tree $T^*$ returned by the \TODO{all variants in Table 1} \algname{} Algorithm~\ref{main_alg}, given a choice of linkage criteria $\interact{}$ and the use (or not) of CannotLinkConstraints. 
Among the algorithms included in the \algname{} framework (see Table \ref{tab:linkage-criteria}), only Mutex Watershed and hierarchical clustering with Average (HC-Avg), Single (HC-Single) and Complete linkage (HC-Complete) define an ultrametric $(V, d_{T^*})$, where $d_{T^*}$ is defined in Eq.~\ref{eq:UM_def} and $T^*$ is the tree returned by the \algname{} Algorithm~\ref{main_alg}.
\end{restatable}
